Need Fulfillment During Intergroup Contact: Three Experience Sampling Studies

### 2023-05-26 Methods-and-Results.Rmd
robustness reports in main text

\subsection{Robustness}
To check for spurious relationships, we test three additional models. These three models assess whether the effect is indeed specific to intergroup contact, whether the need fulillment effect is a affected by whether the contact was planned or accidental, and whether the motivational mechanism also holds direct well-being benefits for the minority members. We describe the full methods and results for the robustness analyses in \appref{app:AppendixRobustness}. We use a two-staged analysis approach for the robustness analyses. We first test the model `globally' --- across the three studies --- in a three level multilevel regression (i.e., measurements nested within participants, and participants nested within studies). Only in a second step do check for study-specific idiosyncrasies.

\subsubsection{Contact specific}
To ensure that the core need fulfillment is outgroup contact specific, we return to the full sample of intensive longitudinal measurements and test whether there is an interaction effect of outgroup contact (vs. no outgroup contact) and core need fulfillment. We expected that the effect of core need fulfillment on outgroup attitudes is specific to outgroup interactions and not merely due to a more need-fulfilled life in general. We find that \hl{...}.

\subsubsection{Interaction intend}
To test whether the need mechanism was affected by whether the interaction was accidental or planned we ran an exploratory moderation analysis using the participants' ratings of how much they perceived the interaction as 'accidental'. It should be noted at this point, that in order to keep the ESM surveys short, we asked our participants to focus on most the important interaction (i.e., "\textit{The following questions will be about the interaction \underline{you consider most significant}.}"''"; emphasis as in original). This was also reflected in a relatively low average and consistent right skew of the 'accidental' item. Nonetheless, there remained a substantial variance in the item and we continued with the moderation analysis. We find that \hl{...}.

\subsubsection{Well-being outcome}
Given the well-established criticism that more positive outgroup attitudes might not always be beneficial for minority group members \citep[e.g.,][]{reimer2023}, we conducted an additional exploratory analysis assessing the effect of need fulfilling outgroup interactions on reported well-being. Experienced well-being is a common indicator of health and an important life quality measurement in itself, especially for migrants or other minority groups \citep[e.g.,][]{bhugra2011}. We find that \hl{...}.
The analysis, thus, adds weight to the importance of considering the need fulfillment experiences when it comes to outgroup interactions.

@@

\subsubsection{Need types}
To ensure that our results are not impacted by differences in the reported goals and motives, we additionally coded the topics we extracted during the topic modeling on two dimensions of how much they reflect a practical and a psychological goal-directedness. We chose practical and psychological needs specifically as our dimensions to account for differences in the types of needs that participants commonly reported. With practical motives we refer to specific, tangible goals or tasks that participants aimed to accomplish during the interaction. These instrumental goals are usually observable, concrete, and often centered on external outcomes, such as acquiring resources, completing tasks, and addressing immediate challenges \citep[][]{}. With psychological motives we refer to underlying motives or desires that are more abstract and relate to personal fulfillment and well-being. In contrast to practical needs, psychological needs delve into the subjective and internal aspects of human experiences. These needs pertain to emotions, social connections, and cognitive processes, reflecting individuals' quest for personal growth, well-being, and thriving in social relationships \citep[][]{}. Note that with this approach any particular motive can include a practical and/or a psychological goal-directedness but can also be classified as not having any goal at all.

For each of the topics, we had identified in the topic modeling analysis, we extracted the top 10 most indicative words as well as the three most representative free text responses. Using a detailed coding protocol for both the practical as well as the psychological motives, two independent coders rated each of the 47 topics. This allowed us to add the two dimensions back to each of the reported interaction motives and assess the distributions and effects of the goal-directed character. We found that \hl{...}.

\subsubsection{Specific psychological needs}
Finally, to ensure that a much simpler model of three fundamental psychological needs might not account for the same effect, we compared our situation core need fulfillment measurement to the three self-determination theory needs (autonomy, relatedness, competence).

>>>>

\subsection{Robustness}
To check for spurious relationships, we test three additional models. These three models assess whether the effect is indeed specific to intergroup contact, whether the need fulfillment effect is a affected by whether the contact was planned or accidental, and whether the motivational mechanism also holds direct well-being benefits for the minority members. We describe the full methods and results for the robustness analyses in \appref{app:AppendixRobustness}. We use a two-staged analysis approach for the robustness analyses. We first test the model `globally' --- across the three studies --- in a three level multilevel regression (i.e., measurements nested within participants, and participants nested within studies). Only in a second step do check for study-specific idiosyncrasies.

\subsubsection{Contact specific}
To ensure that the core need fulfillment is outgroup contact specific, we return to the full sample of intensive longitudinal measurements and test whether there is an interaction effect of outgroup contact (vs. no outgroup contact) and core need fulfillment. We expected that the effect of core need fulfillment on outgroup attitudes is specific to outgroup interactions and not merely due to a more need-fulfilled life in general. Both in the global test, as well as in the individual studies, we consistently find a significant and meaningful interaction effect of outgroup interaction and key need fulfillment, indicating that key need fulfillment is specifically a powerful predictor of outgroup attitudes during intergroup contacts. When assessing the individual studies independently, we additionally find a strong main effect of contact as well as a smaller main effect of key need fulfillment, qualifying the contact-specific relationship (see \appref{app:AppendixRobustness} for full results). In sum, we thus find that at the effect of key need fulfillment on outgroup attitudes is particularly important for outgroup interactions (rather than only need fulfillment in general).

\subsubsection{Interaction intend}
To test whether the need mechanism was affected by whether the interaction was accidental or planned we ran an exploratory moderation analysis using the participants' ratings of how much they perceived the interaction as 'accidental'. It should be noted that, in order to keep the ESM surveys short, we asked our participants to focus on most the important interaction (i.e., "\textit{The following questions will be about the interaction \underline{you consider most significant}.}"; emphasis as in original). This was also reflected in a relatively low average and consistent right skew of the 'accidental' item. Nonetheless, there remained a substantial variance in the item and we continued with the moderation analysis. We found that both in the overall test as in all studies individually key need fulfillment remained a strong predictor of outgroup attitudes, even when accounting for differences in whether the interaction was accidental (rather than planned). Moreover, in none of the analysis did we see a moderation effect of interaction intend nor did we find a main effect of interaction intend. There is, thus, consistent evidence that the need fulfillment mechanism was not meaningfully different for more accidental interactions.

\subsubsection{Well-being outcome}
Given the well-established criticism that more positive outgroup attitudes might not always be beneficial for minority group members \citep[e.g.,][]{reimer2023}, we conducted an additional exploratory analysis assessing the effect of need fulfilling outgroup interactions on reported well-being. Experienced well-being is a common indicator of health and an important life quality measurement in itself, especially for migrants or other minority groups \citep[e.g.,][]{bhugra2011}. We find that the results with well-being as the outcome variable mirror our main results in effect size and interpretation. The results are consistent across and within studies and, thus, add weight to the importance of considering the need fulfillment experiences when it comes to outgroup interactions.

@@

\subsubsection{Need types}
To ensure that our results are not impacted by differences in the reported goals and motives, we additionally coded the topics we extracted during the topic modeling on two dimensions of how much they reflect a practical and a psychological goal-directedness. We chose practical and psychological needs specifically as our dimensions to account for differences in the types of needs that participants commonly reported. With practical motives we refer to specific, tangible goals or tasks that participants aimed to accomplish during the interaction. These instrumental goals are usually observable, concrete, and often centered on external outcomes, such as acquiring resources, completing tasks, and addressing immediate challenges \citep[e.g.,][]{oduntan2019}. With psychological motives we refer to underlying motives or desires that are more abstract and relate to personal fulfillment and well-being. In contrast to practical needs, psychological needs delve into the subjective and internal aspects of human experiences. These needs pertain to emotions, social connections, and cognitive processes, reflecting individuals' quest for personal growth, well-being, and thriving in social relationships \citep[e.g.,][]{dweck2017}. Note that with this approach any particular motive can include a practical and/or a psychological goal-directedness but can also be classified as not having any goal at all. We found high inter-rater reliability for the two independent coders, who coded all 47 topics. The codings revealed considerable variance, which allowed us to add the two dimensions back to each of the reported interaction motives and assess the effects of the goal-directed character.

We found that neither in the overall analysis nor within any of the individual studies did different types of motives impact the positive effect of situational need fulfillment on outgroup attitudes. Situational need fulfillment ratings remained the only reliable predictor of outgroup attitudes, when controlling for practical and psychological goal-directedness as well as their interaction effect with the need fulfillment itself. This result underscores the importance of the experience of perceived need fulfillment (i.e., the perception that one got what one needed), rather than the type of need or the content of the need (i.e., the exact motive).

\subsubsection{Specific psychological needs}
Finally, to ensure that a much simpler model of three fundamental psychological needs might not account for the same effect, we compared our situation core need fulfillment measurement to the three self-determination theory needs (autonomy, relatedness, competence). Within the overall model we found that across the studies core need fulfillment remained a strong predictor of outgroup attitudes, even after controlling for the three self-determination theory need. Additionally, within this overall analysis, none of the self-determination theory needs independently predicted outgroup attitudes to a statistically significant extent (despite similar effect sizes of relatedness). When looking at the individual studies, we again saw that core need fulfillment remained a consistent predictor of outgroup attitudes. However, across all three studies the fulfillment of relatedness motives also emerged as a consistent predictor of outgroup attitudes. Additionally, in the larger studies 2 and 3 competence fulfillment was also related to more positive outgroup attitudes. None of the autonomy fulfillment effects reached statistical significance. In short, find that across our samples, relatedness fulfillment (and to a smaller extend competence fulfillment) are instrumental in understanding when an outgroup contact leads to more positive outgroup attitudes. Importantly, even when considering these effects situational core need fulfillment remains a strong and consistent predictor of outgroup attitudes.

### 2023-05-26 Methods-and-Results.tex
robustness reports in main text

To check for spurious relationships, we test three additional models.
These three models assess whether the effect is indeed specific to
intergroup contact, whether the need fulillment effect is a affected by
whether the contact was planned or accidental, and whether the
motivational mechanism also holds direct well-being benefits for the
minority members. We describe the full methods and results for the
robustness analyses in \appref{app:AppendixRobustness}. We use a
two-staged analysis approach for the robustness analyses. We first test
the model `globally' --- across the three studies --- in a three level
multilevel regression (i.e., measurements nested within participants,
and participants nested within studies). Only in a second step do check
for study-specific idiosyncrasies.

@@

To ensure that the core need fulfillment is outgroup contact specific,
we return to the full sample of intensive longitudinal measurements and
test whether there is an interaction effect of outgroup contact (vs.~no
outgroup contact) and core need fulfillment. We expected that the effect
of core need fulfillment on outgroup attitudes is specific to outgroup
interactions and not merely due to a more need-fulfilled life in
general. We find that \hl{...}.

@@

To test whether the need mechanism was affected by whether the
interaction was accidental or planned we ran an exploratory moderation
analysis using the participants' ratings of how much they perceived the
interaction as `accidental'. It should be noted at this point, that in
order to keep the ESM surveys short, we asked our participants to focus
on most the important interaction (i.e.,
``\textit{The following questions will be about the interaction \underline{you consider most significant}.}'''```;
emphasis as in original). This was also reflected in a relatively low
average and consistent right skew of the 'accidental' item. Nonetheless,
there remained a substantial variance in the item and we continued with
the moderation analysis. We find that \hl{...}.

@@

Given the well-established criticism that more positive outgroup
attitudes might not always be beneficial for minority group members
\citep[e.g.,][]{reimer2023}, we conducted an additional exploratory
analysis assessing the effect of need fulfilling outgroup interactions
on reported well-being. Experienced well-being is a common indicator of
health and an important life quality measurement in itself, especially
for migrants or other minority groups \citep[e.g.,][]{bhugra2011}. We
find that \hl{...}. The analysis, thus, adds weight to the importance of
considering the need fulfillment experiences when it comes to outgroup
interactions.

@@

To ensure that our results are not impacted by differences in the
reported goals and motives, we additionally coded the topics we
extracted during the topic modeling on two dimensions of how much they
reflect a practical and a psychological goal-directedness. We chose
practical and psychological needs specifically as our dimensions to
account for differences in the types of needs that participants commonly
reported. With practical motives we refer to specific, tangible goals or
tasks that participants aimed to accomplish during the interaction.
These instrumental goals are usually observable, concrete, and often
centered on external outcomes, such as acquiring resources, completing
tasks, and addressing immediate challenges \citep[][]{}. With
psychological motives we refer to underlying motives or desires that are
more abstract and relate to personal fulfillment and well-being. In
contrast to practical needs, psychological needs delve into the
subjective and internal aspects of human experiences. These needs
pertain to emotions, social connections, and cognitive processes,
reflecting individuals' quest for personal growth, well-being, and
thriving in social relationships \citep[][]{}. Note that with this
approach any particular motive can include a practical and/or a
psychological goal-directedness but can also be classified as not having
any goal at all.

For each of the topics, we had identified in the topic modeling
analysis, we extracted the top 10 most indicative words as well as the
three most representative free text responses. Using a detailed coding
protocol for both the practical as well as the psychological motives,
two independent coders rated each of the 47 topics. This allowed us to
add the two dimensions back to each of the reported interaction motives
and assess the distributions and effects of the goal-directed character.
We found that \hl{...}.

@@

Finally, to ensure that a much simpler model of three fundamental
psychological needs might not account for the same effect, we compared
our situation core need fulfillment measurement to the three
self-determination theory needs (autonomy, relatedness, competence).

>>>>

To check for spurious relationships, we test three additional models.
These three models assess whether the effect is indeed specific to
intergroup contact, whether the need fulfillment effect is a affected by
whether the contact was planned or accidental, and whether the
motivational mechanism also holds direct well-being benefits for the
minority members. We describe the full methods and results for the
robustness analyses in \appref{app:AppendixRobustness}. We use a
two-staged analysis approach for the robustness analyses. We first test
the model `globally' --- across the three studies --- in a three level
multilevel regression (i.e., measurements nested within participants,
and participants nested within studies). Only in a second step do check
for study-specific idiosyncrasies.

@@

To ensure that the core need fulfillment is outgroup contact specific,
we return to the full sample of intensive longitudinal measurements and
test whether there is an interaction effect of outgroup contact (vs.~no
outgroup contact) and core need fulfillment. We expected that the effect
of core need fulfillment on outgroup attitudes is specific to outgroup
interactions and not merely due to a more need-fulfilled life in
general. Both in the global test, as well as in the individual studies,
we consistently find a significant and meaningful interaction effect of
outgroup interaction and key need fulfillment, indicating that key need
fulfillment is specifically a powerful predictor of outgroup attitudes
during intergroup contacts. When assessing the individual studies
independently, we additionally find a strong main effect of contact as
well as a smaller main effect of key need fulfillment, qualifying the
contact-specific relationship (see \appref{app:AppendixRobustness} for
full results). In sum, we thus find that at the effect of key need
fulfillment on outgroup attitudes is particularly important for outgroup
interactions (rather than only need fulfillment in general).

@@

To test whether the need mechanism was affected by whether the
interaction was accidental or planned we ran an exploratory moderation
analysis using the participants' ratings of how much they perceived the
interaction as `accidental'. It should be noted that, in order to keep
the ESM surveys short, we asked our participants to focus on most the
important interaction (i.e.,
``\textit{The following questions will be about the interaction \underline{you consider most significant}.}'';
emphasis as in original). This was also reflected in a relatively low
average and consistent right skew of the `accidental' item. Nonetheless,
there remained a substantial variance in the item and we continued with
the moderation analysis. We found that both in the overall test as in
all studies individually key need fulfillment remained a strong
predictor of outgroup attitudes, even when accounting for differences in
whether the interaction was accidental (rather than planned). Moreover,
in none of the analysis did we see a moderation effect of interaction
intend nor did we find a main effect of interaction intend. There is,
thus, consistent evidence that the need fulfillment mechanism was not
meaningfully different for more accidental interactions.

@@

Given the well-established criticism that more positive outgroup
attitudes might not always be beneficial for minority group members
\citep[e.g.,][]{reimer2023}, we conducted an additional exploratory
analysis assessing the effect of need fulfilling outgroup interactions
on reported well-being. Experienced well-being is a common indicator of
health and an important life quality measurement in itself, especially
for migrants or other minority groups \citep[e.g.,][]{bhugra2011}. We
find that the results with well-being as the outcome variable mirror our
main results in effect size and interpretation. The results are
consistent across and within studies and, thus, add weight to the
importance of considering the need fulfillment experiences when it comes
to outgroup interactions.

@@

To ensure that our results are not impacted by differences in the
reported goals and motives, we additionally coded the topics we
extracted during the topic modeling on two dimensions of how much they
reflect a practical and a psychological goal-directedness. We chose
practical and psychological needs specifically as our dimensions to
account for differences in the types of needs that participants commonly
reported. With practical motives we refer to specific, tangible goals or
tasks that participants aimed to accomplish during the interaction.
These instrumental goals are usually observable, concrete, and often
centered on external outcomes, such as acquiring resources, completing
tasks, and addressing immediate challenges \citep[e.g.,][]{oduntan2019}.
With psychological motives we refer to underlying motives or desires
that are more abstract and relate to personal fulfillment and
well-being. In contrast to practical needs, psychological needs delve
into the subjective and internal aspects of human experiences. These
needs pertain to emotions, social connections, and cognitive processes,
reflecting individuals' quest for personal growth, well-being, and
thriving in social relationships \citep[e.g.,][]{dweck2017}. Note that
with this approach any particular motive can include a practical and/or
a psychological goal-directedness but can also be classified as not
having any goal at all. We found high inter-rater reliability for the
two independent coders, who coded all 47 topics. The codings revealed
considerable variance, which allowed us to add the two dimensions back
to each of the reported interaction motives and assess the effects of
the goal-directed character.

We found that neither in the overall analysis nor within any of the
individual studies did different types of motives impact the positive
effect of situational need fulfillment on outgroup attitudes.
Situational need fulfillment ratings remained the only reliable
predictor of outgroup attitudes, when controlling for practical and
psychological goal-directedness as well as their interaction effect with
the need fulfillment itself. This result underscores the importance of
the experience of perceived need fulfillment (i.e., the perception that
one got what one needed), rather than the type of need or the content of
the need (i.e., the exact motive).

@@

Finally, to ensure that a much simpler model of three fundamental
psychological needs might not account for the same effect, we compared
our situation core need fulfillment measurement to the three
self-determination theory needs (autonomy, relatedness, competence).
Within the overall model we found that across the studies core need
fulfillment remained a strong predictor of outgroup attitudes, even
after controlling for the three self-determination theory need.
Additionally, within this overall analysis, none of the
self-determination theory needs independently predicted outgroup
attitudes to a statistically significant extent (despite similar effect
sizes of relatedness). When looking at the individual studies, we again
saw that core need fulfillment remained a consistent predictor of
outgroup attitudes. However, across all three studies the fulfillment of
relatedness motives also emerged as a consistent predictor of outgroup
attitudes. Additionally, in the larger studies 2 and 3 competence
fulfillment was also related to more positive outgroup attitudes. None
of the autonomy fulfillment effects reached statistical significance. In
short, find that across our samples, relatedness fulfillment (and to a
smaller extend competence fulfillment) are instrumental in understanding
when an outgroup contact leads to more positive outgroup attitudes.
Importantly, even when considering these effects situational core need
fulfillment remains a strong and consistent predictor of outgroup
attitudes.
